preciseTAD: A machine learning framework for precise TAD boundary prediction at base pair resolution

### 2020-07-06 Manuscript/manuscript.Rmd
WIP on results

Cell line-specific genomic annotations were used to build the feature space for our boundary prediction framework. Bins were annotated by one of either the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), the average signal strength of the corresponding annotation (*Peak Signal Strength*), or the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*). A $log_2$-transformation [??? Shouldn't it be log2 + 1?] of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). It should be noted that models built using a *Peak Signal Strength* feature space were only composed of histone modifications and transcription factor binding sites. That is, they did not include any BroadHMM chromatin segmentations.

@@

Many genomic annotations, notably architectural proteins, tend to exhibit an extensive pattern of colocalization (correlation) [@van2014insulator]. To suitably reduce the feature space and improve computational efficiency, while maintaining optimal performance, we implemented recursive feature elimination (RFE). In doing so, we estimated the near optimal number of necessary features to include in the best fitting RF model. The best performing subset was identified, ranging from 2 to the maximum number of features by powers of 2. We then aggregated the predictive importance of the union of the optimal set of features across chromosomes using the mean decrease in node impurity among permuted features in out-of-bag samples to determine the most common and top ranked annotations for predicting boundary regions.

@@

To investigate whether we could alleviate resolution limitations of conventional TAD and chromatin loop calling tools, we developed *preciseTAD*, an algorithm that leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base pair (bp) resolution. To demonstrate, we implemented an optimally built RF model using the framework described above. That is, a random forest model was built on the optimal combination of data resolution (binning), predictor type (feature space), re-sampling technique (class balancing correction), and top ranked annotations (most informative genomic elements).

@@

Our framework is resolution-specific and uses called TAD or loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines obtained from GEO GSE63525 (Supplementary Table 1). TADs were subsequently called using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loop boundaries, we opted to use boundaries obtained by the PEAKACHU prediction tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? TAD and loop boundaries should be distinguished, to describe when we have all results]

Boundary regions were defined as regions containing a called boundary ($Y = 1$), while non-boundary regions were defined as genomic bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of TADs, TAD boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 3). The number of non-boundary regions highly outnumbered boundary regions, with results consistent for loop boundaries as well (Supplementary Table 4). Such disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects model performance [@prati2011survey; @wei2013role]. To address class imbalance, we evaluated the effect of three re-sampling techniques on prediction performance. *Random over-sampling* (ROS) was defined as sampling with replacement from boundary regions. *Random under-sampling* (RUS) was defined as sampling with replacement from non-boundary regions. Lastly, we considered *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling [@chawla2002smote] (see Methods). These strategies allowed us to select the best performing model.

A total of 78 cell line specific functional genomic annotations were used to build the feature space of downstream random forest models. These annotations included BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS, Supplementary Table 2). Four feature engineering procedures were developed for each genomic annotation (Figure 1B, see Methods) with the goal to improve boundary predictions by accounting for signal strength association (Signal), direct (OC), proportional (OP), and spatial (Distance) relationships between TAD boundaries and genomic annotations.

![**Illustrating the resolution-specific data construction and feature engineering used for downstream random forest modeling.** (A) The linear genome was subsequently binned into resolution-specific intervals. The response vector **Y** was defined as a genomic bin that either overlapped with a TAD boundary ($Y = 1$) or did not ($Y = 0$). (B) Four types of associations between bins and genomic annotations (green shapes) were then used to build the feature space including _Peak Signal_ (Signal), _Overlap Counts_ (OC), _Overlap Percent_ (OP), and ($log_{2}$) _Distance_.](Figures/figure1_model_construction_and_predictor_types.png)

In summary, we considered combinations of data from a particular cell line *L = {GM12878, K562}*, at a given resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, with a specific engineered feature space *P = {Signal, OC, OP, Distance}*, and re-sampling technique *S = {None, RUS, ROS, SMOTE}* (Figure 2). Once the model inputs were established, a random forest classifier was trained on $n-1$ autosomal chromosomes, while reserving the $i^{th}$ chromosome for testing. Three-fold cross-validation was used to tune the *mtry* hyperparameter, while *ntree* and *nodesize* were fixed to default values of 500 and 1 respectively. Model inputs were evaluated by aggregating the mean balanced accuracy (BA) across each holdout chromosome (See Methods).

![**A diagram of the machine learning framework detailing the steps for combining different inputs used in the random forest model building pipeline.** Step 1 is to establish the chromosome-specific rowise (genomic bins) and columnwise (genomic annotations) dimensions of the data matrix, **A**, by combining a resolution (5 kb, 10 kb, 25 kb, 50 kb, or 100 kb) and feature engineering procedure (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)) for the $n-1$ autosomal chromosomes. Step 2 is to reserve the binned data matrix for the holdout chromosome $i$ as the test data. Step 3 is to apply a re-sampling technique to the training set (no re-sampling, random over-sampling, random under-sampling, or SMOTE re-sampling). Step 4 is to train the random forest model and perform 3-fold cross-validation to tune for the *mtry* parameter. Finally, step 5 is to validate the model on the separate test data and evaluate model performance using balanced accuracy (BA).](Figures/figure2_ensemble_framework.png)

@@

Additionally, we found that at higher resolutions, random forest classifiers built using a distance-type feature space yielded substantially higher balanced accuracies when compared to the peak signal stength, overlap count, and overlap percent predictor types. As with comparing class balancing techniques, this improvement was less evident at lower resolutions. These results were found to be consistent for the K562 cell line (Supplementary Figure 2). Random forest models built on PEAKACHU loop boundaries with SMOTE resampling only marginally outperformed RUS for both GM12878 and K562 cell lines (Supplementary Figure 3). Five kilobase resolution genomic bins was found to result in optimal prediction for both TAD and loop boundaries, on both cell lines. Our results indicate that random under-sampling, distance-type predictors, and Hi-C data at 5kb resolution are the optimal set of data level characteristics for both TAD and chromatin loop boundary region prediction.

@@

![**Assessing top ranked transcription factors in TAD and loop boundary region prediction for GM12878.** (A) Barplots comparing performances of TAD (ARROWHEAD) and loop (PEAKACHU) boundary region prediction models using distance-type specific classes of functional genomic elements as the feature space including histone modifications (HM), cis-regulated elements (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to reduce the number of TFBS to include in the optimally built random forest model. Error bars represent standard deviation from the cross-validated balance accuracy across each autosomal chromosome. (C) Clustered heatmap of predictive importances for the union of the top 8 most predictive TFBS across chromosomes as determined from performing RFE.](Figures/figure4_annots_specifc_rfe_importances_arrowhead_peakachu.png)

@@

Here we present *preciseTAD*, a random forest-based approach coupled with spatial clustering (DBSCAN) and scalable partition around medoids (CLARA) that allows for more precise prediction of TAD or loop boundaries. *preciseTAD* leverages a random forest (RF) classification algorithm built on ground truth TAD(loop) boundaries obtained from a TAD(loop) calling tool and ChIP-seq defined genomic annotations as the feature space. We first optimized our RF model by creating a machine learning framework that systematically compared the performances of different combinations of genome binning (resolution), feature engineering procedures, and re-sampling technique. In doing so, we first introduced a new strategy for establishing boundary region prediction data referred to as *shifted binning*, in which the rows of the data matrix were controlled by the data resolution (either 5kb, 10kb, 25kb, 50kb, and 100kb) used to obtain ground truth boundaries, while assuring ground truth bounderies were located central to any given genomic bin. Furthermore, we introduced a new procedure for drawing spatial associations between boundary regions and genomic annotations, referred to as *distance*-type predictors, defined as as the $log_{2}$ genomic distance in base pairs from each genomic bin the nearest ChIP-seq peak region. We benchmarked *preciseTAD* against two boundary calling tools, ARROWHEAD [@durand2016juicebox], an established TAD-caller, and PEAKACHU [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. For a more in depth study, we also applied *preciseTAD* on cell line-specific boundaries and ChIP-seq data from both the GM12878 and K562 cell lines.

@@

**Supplementary Table 2.** A complete list of the genomic annotations used to build the feature space for all downstream models. The GRCh37/hg19 human genome assembly was used. "Genomic Class" - broad category of genomic features, "Element" - names of genomic features, "Cell Type-Specific Source" - download URL specific to the cell line (not all annotations were provided by the same institutions).

@@

**Supplementary Figure 4. Assessing top ranked transcription factors in TAD and loop boundary prediction for K562.** (A) Barplots comparing performances of TAD (ARROWHEAD) and loop (PEAKACHU) boundary prediction models using specific classes of functional genomic elements as the feature space including histone modifications (HM), cis-regulated elements (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to reduce the number of TFBS to include in the optimally built random forest model. Error bars represent standard deviation from the cross-validated accuracy across each chromosome. (C) Clustered heatmap of predictive importances for the union of the top 8 most predictive TFBS across chromosomes as determined from performing RFE.

>>>>

Cell line-specific genomic annotations were used to build the predictor space for our boundary prediction framework. Bins were annotated by one of either the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), the average signal strength of the corresponding annotation (*Peak Signal Strength*), or the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*). A $log_2$-transformation [??? Shouldn't it be log2 + 1?] of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). Models built using a *Peak Signal Strength* predictor space were only composed of histone modifications and transcription factor binding sites because BroadHMM chromatin segmentations lack signal values.

@@

Many genomic annotations, notably architectural proteins, tend to exhibit an extensive pattern of colocalization (correlation) [@van2014insulator]. To suitably reduce the predictor space and improve computational efficiency, while maintaining optimal performance, we implemented recursive feature elimination (RFE). In doing so, we estimated the near optimal number of necessary features to include in the best fitting RF model. The best performing subset was identified, ranging from 2 to the maximum number of features by powers of 2. We then aggregated the predictive importance of the union of the optimal set of features across chromosomes using the mean decrease in node impurity among permuted features in out-of-bag samples to determine the most common and top ranked annotations for predicting boundary regions.

@@

To investigate whether we could alleviate resolution limitations of conventional TAD and chromatin loop calling tools, we developed *preciseTAD*, an algorithm that leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base pair (bp) resolution. To demonstrate, we implemented an optimally built RF model using the framework described above. That is, a random forest model was built on the optimal combination of data resolution (binning), predictor type (predictor space), re-sampling technique (class balancing correction), and top ranked annotations (most informative genomic elements).

@@

Our framework is resolution-specific and uses called TAD or loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines obtained from GEO GSE63525 (Supplementary Table 1) [??? Update table and methods with relevant data only, not individual samples]. TADs were subsequently called using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loop boundaries, we opted to use boundaries obtained by the PEAKACHU prediction tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? Is there other resolution? ??? TAD and loop boundaries should be distinguished, to describe when we have all results] [??? Define that ARROWHEAD data constitute the main results]

Boundary regions were defined as regions containing a called boundary ($Y = 1$), while non-boundary regions were defined as genomic bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of TADs, TAD boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 3 [??? I can see numbers for K562 (?) only? There's a discrepancy with legend.]). The number of non-boundary regions highly outnumbered boundary regions, with results consistent for loop boundaries as well (Supplementary Table 4). Such disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address class imbalance, we evaluated the effect of three re-sampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority cluss (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

A total of 78 [??? 77 in the table?] cell-line-specific functional genomic annotations were used to build the predictor space. These included BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS, Supplementary Table 2). Four feature engineering procedures were developed to quantify the association between genomic annotations and bins (Figure 1B). These include signal strength association (Signal), direct (OC), proportional (OP), and spatial (Distance) relationships (see Methods).

![**Resolution-specific data construction and feature engineering used for downstream random forest modeling.** (A) The linear genome was binned into non-overlapping resolution-specific intervals (see Methods). The response vector **Y** was defined as 1/0 if a genomic bin overlapped/did not overlap with a TAD boundary, respectively. (B) Four types of associations between bins and genomic annotations (green shapes) were considered to build the predictor space including _Average Peak Signal_ (Signal), _Overlap Counts_ (OC), _Overlap Percent_ (OP), and ($log_{2}$) _Distance_.](Figures/figure1_model_construction_and_predictor_types.png)

In summary, we considered combinations of data from two cell line *L = {GM12878, K562}*, five resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, four types of predictor space *P = {Signal, OC, OP, Distance}*, and three re-sampling technique *S = {None, RUS, ROS, SMOTE}* (Figure 2). Once the model inputs were established, a random forest classifier was trained on $n-1$ autosomal chromosomes, while reserving the $i^{th}$ chromosome for testing. Three-fold cross-validation was used to tune the *mtry* hyperparameter, while *ntree* and *nodesize* were fixed to default values of 500 and 1 respectively. Model performance was evaluated by aggregating the mean balanced accuracy (BA) across each holdout chromosome (see Methods). These strategies allowed us to select the best performing model.

![**A diagram of the machine learning framework.** Step 1 defines the chromosome-specific predictor-response matrix **A** of $N$ genomic bins (rows), $p$ genomic annotations, and one response vector $Y$ ($p+1$ columns). Resolution-specific data (5 kb, 10 kb, 25 kb, 50 kb, or 100 kb) and feature engineering procedure (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)) for the $n-1$ autosomal chromosomes. Step 2 is to reserve the predictor-response matrix for the holdout chromosome $i$ as the test data. Step 3 is to apply a re-sampling technique to the training set (no re-sampling, random over-sampling, random under-sampling, or SMOTE re-sampling). Step 4 is to train the random forest model and perform 3-fold cross-validation to tune for the *mtry* parameter. Finally, step 5 is to validate the model on the separate test data and evaluate model performance using balanced accuracy (BA).](Figures/figure2_ensemble_framework.png) [??? DEfine all indexes, what is N, p+1, i?]

@@

Additionally, we found that at higher resolutions, random forest classifiers built using a distance-type predictor space yielded substantially higher balanced accuracies when compared to the peak signal stength, overlap count, and overlap percent predictor types. As with comparing class balancing techniques, this improvement was less evident at lower resolutions. These results were found to be consistent for the K562 cell line (Supplementary Figure 2). Random forest models built on PEAKACHU loop boundaries with SMOTE resampling only marginally outperformed RUS for both GM12878 and K562 cell lines (Supplementary Figure 3). Five kilobase resolution genomic bins was found to result in optimal prediction for both TAD and loop boundaries, on both cell lines. Our results indicate that random under-sampling, distance-type predictors, and Hi-C data at 5kb resolution are the optimal set of data level characteristics for both TAD and chromatin loop boundary region prediction.

@@

![**Assessing top ranked transcription factors in TAD and loop boundary region prediction for GM12878.** (A) Barplots comparing performances of TAD (ARROWHEAD) and loop (PEAKACHU) boundary region prediction models using distance-type specific classes of functional genomic elements as the predictor space including histone modifications (HM), cis-regulated elements (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to reduce the number of TFBS to include in the optimally built random forest model. Error bars represent standard deviation from the cross-validated balance accuracy across each autosomal chromosome. (C) Clustered heatmap of predictive importances for the union of the top 8 most predictive TFBS across chromosomes as determined from performing RFE.](Figures/figure4_annots_specifc_rfe_importances_arrowhead_peakachu.png)

@@

Here we present *preciseTAD*, a random forest-based approach coupled with spatial clustering (DBSCAN) and scalable partition around medoids (CLARA) that allows for more precise prediction of TAD or loop boundaries. *preciseTAD* leverages a random forest (RF) classification algorithm built on ground truth TAD(loop) boundaries obtained from a TAD(loop) calling tool and ChIP-seq defined genomic annotations as the predictor space. We first optimized our RF model by creating a machine learning framework that systematically compared the performances of different combinations of genome binning (resolution), feature engineering procedures, and re-sampling technique. In doing so, we first introduced a new strategy for establishing boundary region prediction data referred to as *shifted binning*, in which the rows of the data matrix were controlled by the data resolution (either 5kb, 10kb, 25kb, 50kb, and 100kb) used to obtain ground truth boundaries, while assuring ground truth bounderies were located central to any given genomic bin. Furthermore, we introduced a new procedure for drawing spatial associations between boundary regions and genomic annotations, referred to as *distance*-type predictors, defined as as the $log_{2}$ genomic distance in base pairs from each genomic bin the nearest ChIP-seq peak region. We benchmarked *preciseTAD* against two boundary calling tools, ARROWHEAD [@durand2016juicebox], an established TAD-caller, and PEAKACHU [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. For a more in depth study, we also applied *preciseTAD* on cell line-specific boundaries and ChIP-seq data from both the GM12878 and K562 cell lines.

@@

**Supplementary Table 2.** A complete list of the genomic annotations used to build the predictor space for all downstream models. The GRCh37/hg19 human genome assembly was used. "Genomic Class" - broad category of genomic features, "Element" - names of genomic features, "Cell Type-Specific Source" - download URL specific to the cell line (not all annotations were provided by the same institutions).

@@

**Supplementary Figure 4. Assessing top ranked transcription factors in TAD and loop boundary prediction for K562.** (A) Barplots comparing performances of TAD (ARROWHEAD) and loop (PEAKACHU) boundary prediction models using specific classes of functional genomic elements as the predictor space including histone modifications (HM), cis-regulated elements (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to reduce the number of TFBS to include in the optimally built random forest model. Error bars represent standard deviation from the cross-validated accuracy across each chromosome. (C) Clustered heatmap of predictive importances for the union of the top 8 most predictive TFBS across chromosomes as determined from performing RFE.
